---
title: "The year in *content*: 2024"
subtitle: "Music, books, and movies"
author: "Juan Tellez"
date: "2023-12-31"
toc: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "85%",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache.extra = 1234
)

options(
  digits = 4,
  width = 300,
  dplyr.summarise.inform = FALSE
)

# Libraries
library(tidyverse)
library(kableExtra)

# Nice ggplot theme
theme_nice = function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = rel(1.25)),
      plot.subtitle = element_text(face = "plain"),
      plot.caption = element_text(face = "plain"),
      axis.title = element_text(face = "bold", size = rel(0.8)),
      axis.title.x = element_text(hjust = 0),
      axis.title.y = element_text(hjust = 1),
      strip.text = element_text(
        face = "bold",
        size = rel(0.8), hjust = 0
      ),
      strip.background = element_rect(fill = "grey90", color = NA),
      legend.title = element_text(size = rel(0.8)),
      legend.text = element_text(size = rel(0.8)),
      legend.position = "bottom",
      legend.justification = "left",
      legend.title.position = "top",
      legend.margin = margin(l = 0, t = 0)
    )
}

theme_set(theme_nice())

# Palette
pal <- MoMAColors::moma.colors("VanGogh")

# kbl wrapper
my_kbl = function(x, ...) {
  kbl(
    x,
    digits = 2,
    booktabs = TRUE,
    align = "c",
    ...
  ) |>
    kable_styling(
      full_width = FALSE,
      bootstrap_options = c("striped", "hover"),
      position = "center"
    )
}

# Load data
music = readr::read_rds(here::here("personal_files/itunes_library.rds"))
books = readr::read_rds(here::here("personal_files/goodreads_clean.rds")) |>
  filter(year(date_read) == 2024)
movies = read_rds(here::here("personal_files/letterboxd_clean.rds")) |>
  filter(year(watched_date) == 2024)
```

# Intro

Here's all the content I consumed this year. The music data comes from Apple Music, the books from Goodreads, and the movies from Letterboxd.

The code I used to clean up the Apple Music data is slightly adapted from Andrew Heiss, which is slightly adapted from somewhere else.

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| echo: true

read_itunes_library <- function(path) {
  raw <- xml2::read_xml(path)

  res <- xml2::as_list(raw)
  res <- purrr::pluck(res, "plist", "dict", "dict")
  res <- res[names(res) != "key"]

  res <-
    tibble::enframe(res) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(value = list(tibble::enframe(value))) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(id = dplyr::row_number()) %>%
    dplyr::select(-name) %>%
    tidyr::unnest(value) %>%
    dplyr::mutate(
      entry_id = (dplyr::row_number() + (dplyr::row_number() %% 2)) / 2
    ) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      value = dplyr::if_else(length(value) == 0L, list(list(NA)), list(value)),
      value = unlist(value)
    ) %>%
    dplyr::ungroup() %>%
    tidyr::pivot_wider(
      id_cols = c(id, entry_id),
      names_from = name,
      values_from = value,
      values_fn = list
    ) %>%
    tidyr::pivot_longer(
      cols = 4:ncol(.),
      names_to = "type",
      values_drop_na = TRUE
    ) %>%
    dplyr::select(-type) %>%
    tidyr::pivot_wider(id_cols = id, names_from = key, values_from = value) %>%
    janitor::clean_names() %>%
    dplyr::select(
      id, track_title = name, artist, album_artist, album,
      genre, total_time, date_added, skip_count, play_count, year
    ) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(dplyr::across(
      everything(),
      ~dplyr::if_else(is.null(.x), list(NA), list(.x))
    )) %>%
    dplyr::mutate(dplyr::across(everything(), unlist)) %>%
    dplyr::mutate(
      date_added = strsplit(date_added, "T"),
      date_added = date_added[1],
      date_added = lubridate::ymd(date_added),
      skip_count = as.numeric(skip_count),
      play_count = as.numeric(play_count),
      total_time = as.numeric(total_time),
      year = as.numeric(year)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(dplyr::desc(play_count))

  res
}
```

# Music

## New albums in my library

First, the **albums I added to my Apple Music library this year**. These are the records I discovered this year, whether they came out this year or earlier. Below are the top five by play count^[One limitation is that, all else equal, albums with more songs will have higher play counts.].

```{r}
music |>
  filter(year(date_added) == 2024) |>
  filter(genre != "Holiday") |>
  group_by(artist, album) |>
  summarise(play_count = sum(play_count), .groups = "drop") |>
  slice_max(order_by = play_count, n = 5) |>
  select(Artist = artist, Album = album) |>
  my_kbl(caption = "Top albums added to my Apple Music library in 2024")
```

- *Adrianne Lenker* is always good, respect for musicians who put out a lot of music. Some great tracks on this one, and I love the versions on the live album.
- *Karen Dalton*, new discovery for me. Odd voice, and I love the bass on some of these songs.
- The *Nap Eyes* record is great. They came to that vineyard near here but I couldn't go see them.

## Favorite records of the year

These are the records that *came out this year* and that I *added to my library this year*.

```{r}
music |>
  filter(album %in% c(
    "Paris 1919 (Deluxe Edition)",
    "I Wanna Run Barefoot Through Your Hair",
    "Magnolia"
  )) |>
  distinct(album, .keep_all = TRUE) |>
  select(Artist = artist, Album = album, Genre = genre) |>
  my_kbl(caption = "Favorite albums of 2024")
```

**For my favorites that came out this year**, I'll pick:

- This great, piano-forward jazz record by *Okonski*.
- The new *Christopher Owens* solo record. Listened to Girls for years, and this one still surprised me.
- *Paris 1919* did not come out this year, obviously, but this deluxe edition did, which gives me an excuse to list one of my favorite records of all time. Cale is a genius, has such a sweet perspective, great Welsh accent.

# Books

## Books I read this year

Here's the books I finished reading this year and my rating:

```{r results='asis'}
last_books <- books %>%
  mutate(`Date Read` = date_read) |>
  filter(!is.na(`Date Read`)) |>
  arrange(desc(`Date Read`)) |>
  mutate(
    formatted_date = format(`Date Read`, "%B %d, %Y"),
    stars = map_chr(`my_rating`, ~ strrep("★", .x))
  )

last_books %>%
  mutate(item = glue::glue(
    "<li>
      <strong>{title}</strong> by {author}<br>
      <em>My Rating:</em> {stars}<br>
      <em>Date Read:</em> {formatted_date}<br>
    </li>"
  )) %>%
  pull(item) %>%
  {cat("<ul>\n", paste(., collapse = "\n"), "\n</ul>")}
```

## Descriptives

How do my ratings compare versus the averages reported by Goodreads?

```{r}
#| code-fold: true
#| fig.height: 6

books |>
  select(title, my_rating, average_rating) |>
  mutate(title = str_remove(title, ":.*$")) |>
  mutate(diff = my_rating - average_rating) |>
  ggplot(aes(y = reorder(title, diff), x = diff, fill = diff > 0)) +
  geom_col(alpha = 0.8) +
  labs(
    title = "Difference between my ratings and the\naverage Goodreads user",
    x = "My rating - average Goodreads user",
    fill = "Books I liked:",
    y = NULL
  ) +
  scale_fill_manual(
    values = pal[c(1, 7)],
    labels = c(
      "...less than the Goodreads average",
      "...more than the Goodreads average"
    )
  ) +
  theme(
    panel.grid.major = element_blank(),
    axis.line = element_blank(),
    legend.position = "top"
  )
```

How long am I spending on each book? Caveat: the dates added / read are approximations in some cases.

```{r}
#| code-fold: true
#| fig.height: 6

books_bell = books %>%
  mutate(title = str_remove(title, ":.*$")) |>
  mutate(title = fct_reorder(title, date_read))

hilite = books_bell |>
  mutate(duration = date_read - date_added) |>
  slice_max(order_by = duration, n = 3) |>
  mutate(title = str_remove(title, ":.*$")) |>
  mutate(title = fct_reorder(title, date_added))

ggplot(filter(books_bell, date_added >= ymd("2023-01-01")), aes(y = title)) +
  geom_segment(
    aes(x = date_added, xend = date_read, yend = title),
    color = "gray70",
    size = 2
  ) +
  geom_point(aes(x = date_added), color = pal[1], size = 4) +
  geom_point(aes(x = date_read), color = pal[7], size = 4) +
  labs(
    title = "Time spent on each book",
    x = "Date",
    y = NULL
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months")
```

## Favorite books

**Favorite books** I read this year:

- *Lituma en los Andes*. A sorta police procedural / mystery novel set in the Peruvian highlands during the time of Shining Path. Incredible atmosphere and treatment of SP as a kind of ghost haunting this town.
- I can't believe I read *Temporada de huracanes* and *2666* in the same year. Both incredible. Bolaño just an unbelievable writer, he can do anything. And *Huracanes* gives us something new: a "Gummo"-like view of a rural Mexican town where deindustrialization has hollowed out the community.

**Surprise read**: *My Year of Rest and Relaxation*. The gender-flipped version of American Psycho. I'm a big Ottessa Moshfegh fan now.

**Worst book**: *Una historia política de Colombia*. Repetitive, not much new to say.

# Movies

You can see all the movies I watched this year at the bottom. Here's my stats for the year:

```{r}
movies |>
  summarise(
    `Number of movies watched` = n(),
    `Average rating` = mean(rating),
    `Highest rated movie` = name[which.max(rating)],
    `Lowest rated movie` = name[which.min(rating)],
    `Cor(rating, release date)` = cor(rating, year)
  ) |>
  mutate(across(is.numeric, ~round(.x, 2))) |>
  mutate(across(everything(), as.character)) |>
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") |>
  my_kbl(caption = "My movie watching stats for 2024")
```

Here's when I watched:

```{r}
#| fig.height: 4

movies |>
  mutate(
    month = month(watched_date, label = TRUE),
    day = day(watched_date)
  ) |>
  count(month, day) |>
  ggplot(aes(x = day, y = month, fill = factor(n))) +
  geom_tile(color = "white") +
  coord_fixed(ratio = 1) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20, 25, 31), expand = c(0, 0)) +
  scale_fill_manual(values = c(pal[5], pal[3])) +
  theme(
    panel.grid = element_blank(),
    panel.grid.major.y = element_line(color = "grey92", linewidth = 0.3),
    axis.title = element_blank(),
    axis.text = element_text(color = "grey30"),
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(5, 5, 5, 5)
  ) +
  labs(x = NULL, y = NULL, title = "Movie watching calendar", fill = NULL)
```

And here's the distribution of ratings, for fun:

```{r}
#| fig.height: 4

movies |>
  ggplot(aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = pal[7], color = "white", alpha = .8) +
  labs(x = "My rating", y = "Number of movies watched") +
  scale_x_continuous(breaks = seq(0, 5, by = 0.5))
```

## Favorite movies

**The two best movies I saw this year** were:

- *Red Rooms*. One of those movies that is completely engrossing and then you take a beat and realize you've just been watching a few actors, some shots in a courtroom, an apartment, etc. No bells and whistles, just great writing and pacing.
- *Possession*. Had heard all the lore and finally watched it at *Dreamland Cinema*. The worst couple of all time.

**The worst movie I saw this year** was *Civil War*. Really it should be *Anyone But You* but that movie isn't putting on airs. I thought *Civil War* was totally undermined by its focus on small potatoes photo-journalism. Who cares?

**The best movie that came out this year** was *Daughters*, the documentary about a prison program that connects inmates with their daughters for a father-daughter dance. Ugly cry kind of film.

## Full list of movies

```{r}
#| code-fold: true

movies |>
  arrange(watched_date) |>
  select(` ` = name, `Release year` = year, `My rating` = rating) |>
  kableExtra::kbl(caption = "Movies I watched this year")
```
