---
title: "Untitled"
format: html
---



## Introduction


When I first learned regression, we would look at a model like this: 

$$Income_i = \beta_1 \times Education_i + \epsilon_i$$

 
We'd then say that for every additional year of Education, Income would be expected to increase by $\beta_1$. We'd then add a control for Age:

$$Income_i = \beta_1 \times Education_i + \beta_2 \times Age_i + \epsilon_i$$

And then something mysterious would happen. We would now say that for every additional year of Education, Income would be expected to increase by $\beta_1$, *controlling for Age*. What did this mean, "controlling for age"? Other, no less mysterious versions of this phrase include "adjusting for age", or "holding age constant". 

Some textbooks would discuss controlling for age in terms of comparisons. Here's Gelman in *Regression and other stories*:

> > We interpret the regression slopes as comparisons of individuals that differ in one predictor while being at the same levels of the other predictors.

Or from *The Effect*:

> If two observations have the same values of the other variables in the model, but one has a value ofð‘‹ that is one unit higher, the observation with the X one unit higher will on average have að‘Œthat is B1 units higher


This was much more intuitive to me. In our case, we'd say that *comparing two people of identical age*, we'd expect someone with an additional year of education to have $\beta_1$ more in earnings. 


But how is regression *doing this*? What about the regression makes this comparison possible? 


Here are two ways of thinking about this that clarified regression *controls*, and what they can and can't tell us. 




```{r}
library(tidyverse)
```


Let's look at some survey data, a sample of the GSS from `socviz`:

```{r}
library(socviz)
gss_sm |> 
  select(id, year, obama, childs, age)
```


## "What's left over"




```{r}
mod1 = lm(childs ~ age, data = gss_sm)
modelsummary::modelsummary(mod1)
```


Let's look at the residuals:

```{r}
augmented_data = broom::augment(mod1) 

augmented_data |> 
  select(.rownames, childs, age, .fitted, .resid) |> 
  sample_n(5) |> 
  kableExtra::kable(digits = 2)
```




```{r}
mod2 = lm(obama ~ age, data = gss_sm)
mod2
```


```{r}
broom::augment(mod2)
```

